Deciding Bitvector Termination with Program Synthesis

### 2014-07-02 papers/termination/paper.tex
Move section 5 around a bit

\section{Termination as Second-Order Satisfaction} \label{sec:second.order}
Program model? Programs: transition systems? Constraint generation? No CFG?

@@

\subsection{The Termination Problem}

@@

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, Inv . \forall x, x' . & I(x) \wedge G(x) \rightarrow Inv(x) ~ \wedge \\
                                 & G(x) \wedge Inv(x) \wedge T(x, x') \rightarrow Inv(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

\begin{figure*}
 \begin{framed}
 \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
  \begin{align*}
\label{def:multi-termination-formula}
  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
 & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
 & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
 & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
 & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
 \end{align*}
\end{definition}

\begin{definition}[ Nested Loops Termination Formula {\bf [NT]}]
\label{def:nested-term-formula}
 \begin{align*}
  \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
                                & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
                                P_2(x'',x''') \wedge 
                                  R(x) > 0 \wedge R(x) > R(x''') 
 \end{align*} 
\end{definition}

@@

\noindent {\bf Unconditional termination.}
We start by considering a program consisting of a single loop ${L= \langle G,T \rangle}$,
where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
state $x'$ is reachable from state $x$ via a single iteration of the loop body.
%where the initial states of the program 
%just before the loop are given as an assertion init(V )
%satisfy $I : X \rightarrow \mathbb{B}$, $G : X \rightarrow \mathbb{B} $ is the loop guard and $T : X \rightarrow X$ is the loop body. 
%We use predicates over program states $x$ and their successors $x'$ to model program
%statements as binary relations over states. For loop $L$, $T(x, x')$
%represents the body of the loop. 
%
%Subsequently, we show how to generalise our formulation of termination and non-termination to handle sequential loops. \\
%
The existence of a ranking function for $L$ is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}.  If this formula is satisfiable,
then there exists a ranking function $R$ that proves loop $L$ terminates.

Our formalism is illustrated by the unconditionally terminating program shown
in Figure~\ref{fig:motivation.a}, which is encoded as:
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}

We will abbreviate this with the notation $G(x) \triangleq x > 0$ and $T(x, x') \triangleq x' = (x - 1) \, \& \, x$.
The corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.\\

\noindent {\bf Conditional termination.}
A slightly more involved case occurs if $L$'s termination depends on the state it begins
executing in, i.e. $L$ is \emph{conditionally terminating}. 
The loop we consider now is $L=\langle I,G,T \rangle$ and has an additional predicate $I(x)$ 
representing the states in which the loop can begin executing.
Conditional termination is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
two witnesses are returned:
\begin{itemize}
\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
guard $G$ is met.
\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.
\end{itemize}

Conditional termination is illustrated by the loop in Figure~\ref{fig:motivation.h}, 
which is encoded as:
\begin{align*}
            I(\langle x, y \rangle) & \triangleq y = 1 \\
            G(\langle x, y \rangle) & \triangleq x > 0 \\
            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y \\
\end{align*}

If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
would lead to a non-terminating execution.

However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
This constitutes a proof that the program as whole terminates, since the loop always begins
executing in a state that guarantees its termination.

\subsection{The Non-Termination Problem}

Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
we can prove this by finding a ranking function and supporting invariant witnessing the satisfiability
of formula {\bf[CT]}.  What then would a proof of non-termination look like?

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
i.e. a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$. 
For our case, given a loop $L = \langle I,G,T \rangle$, the notion of open recurrent set is encoded by Definition~\ref{def:nonterm-formula}.

@@

In order to formulate the non-termination problem inside the synthesis fragment, 
we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
As opposed to a closed recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
As our formulation is with respect to a given loop $L=\langle I,G,t \rangle$, we require the  
recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
infinite execution to happen.

Given that our program's state space is finite, a transition relation induces a infinite execution if and only if some state is 
visited infinitely often, or equivalently if the state is visitied twice in the course of the execution.
This gives us an alternative way of expression a non-termination proof: we can can try to find a witness $x$ to the satisfiability
of the formula
$$ \exists x . T^+(x, x)$$
However, deciding satisfiability of this formula directly requires a logic that includes a transitive closure operator, which
ours does not have.\\

@@

>>>>

\section{Termination as Second-Order Satisfaction} \label{sec:second.order}
%Program model? Programs: transition systems? Constraint generation? No CFG?

@@

\subsection{An Isolated Loop}

We will begin our discussion by showing how to encode the (non-)termination of a single loop
in the synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T \rangle}$,
where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
state $x'$ is reachable from state $x$ via a single iteration of the loop body.
For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

@@

@@

\noindent {\bf Unconditional termination.}
To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
state it starts in), we can find a ranking function for $L$.
The existence of a ranking function is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}, so a witness to the satisfiability
of formula {\bf [UT]} is also a ranking function and thus a proof of $L$'s unconditional termination.

Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.

\noindent{\bf Non-termination.}
Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
we can prove this by finding a ranking function and supporting invariant witnessing the satisfiability
of formula {\bf[CT]}.  What then would a proof of non-termination look like?

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
i.e. a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$. 
For our case, given a loop $L = \langle I,G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.

@@

In order to formulate the non-termination problem inside the synthesis fragment, 
we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
As opposed to a closed recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
satisfiability of formula {\bf [CNT]}. %Definition~\ref{def:deterministic-nonterm-formula}.
%As our formulation is with respect to a given loop $L=\langle I,G,T \rangle$, we require the  
%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
%infinite execution to happen.

Given that our program's state space is finite, a transition relation induces a infinite execution if and only if some state is 
visited infinitely often, or equivalently the state is visited twice in the course of the execution.
This gives us an alternative way of expression a non-termination proof: we can can try to find a witness $x$ to the satisfiability
of the formula
$$ \exists x . T^+(x, x)$$
However, deciding satisfiability of this formula directly requires a logic that includes a transitive closure operator, which
ours does not have.\\

@@

\subsection{Composing a Loop with its Environment}

\begin{figure*}
 \begin{framed}

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

 \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
  \begin{align*}
\label{def:multi-termination-formula}
  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
 & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
 & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
 & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
 & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
 \end{align*}
\end{definition}

\begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
\label{def:nested-term-formula}
 \begin{align*}
  \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
                                & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
                                P_2(x'',x''') \wedge 
                                  R(x) > 0 \wedge R(x) > R(x''') 
 \end{align*} 
\end{definition}

Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
the loop may not terminate if started in some particular state, but the rest of the program
ensures that such a state cannot be reached on entry to the loop.  The program as a whole
terminates, but if the loop were considered in isolation we would not be able to prove that
it terminates.  We will now consider the problem of encoding the loop's environment as
a synthesis formula.

\noindent {\bf Conditional termination.}
If $L$'s termination depends on the state it begins
executing in, we say that $L$ is \emph{conditionally terminating}. 
The loop we consider now is $L=\langle I,G,T \rangle$; it has an additional predicate $I(x)$ 
representing the states in which the loop can begin executing.
Conditional termination is equivalent to the
formula of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
two witnesses are returned:
\begin{itemize}
\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
guard $G$ is met.
\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.
\end{itemize}

Conditional termination is illustrated by the program in Figure~\ref{fig:motivation.h},
which is encoded as:
\begin{align*}
            I(\langle x, y \rangle) & \triangleq y = 1 \\
            G(\langle x, y \rangle) & \triangleq x > 0 \\
            T(\langle x, y \rangle, \langle x', y' \rangle) & \triangleq x' = x - y \wedge y' = y 
\end{align*}
If the initial states $I$ are ignored this loop cannot be shown to terminate, since any state with $y = 0$ and $x > 0$
would lead to a non-terminating execution.

However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
$R(\langle x,y\rangle ) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
This constitutes a proof that the program as a whole terminates, since the loop always begins
executing in a state that guarantees its termination.
